Slight Positive Correlation Between Gendered and Racialized Police Strip Searches in the Toronto Municipal, 2020-2021

### 2024-01-24 outputs/paper/paper.qmd
update and refine data cleaning and results

---
title: "Slight Positive Correlation Between Gendered and Racialized Police Strip Searches in the Toronto Municipal, 2020-2021"
title-block-banner: true
abstract: "Police strip searche has been a topic of great contention during the COVID-19 pandemic. The question commonly raised was how certain demographics are more likely to undergo arrests and strip searches than others, and why has this skewed power dynamics been maintained for so long. Data then was gathered from Open Data Toronto to assess the correlation between strip searches and the racial and gender identities of arrested 'suspects' in the Toronto Municipal from 2020 to 2021. This analysis would enable us to rethink the modern form of policing, how it transpires in times of crisis and critically affects people of marginalized identities. More on RESULTS" 
thanks: "Code and data are available at: https://github.com/ponolite/Police_Arrests_and_Strip_Searches/tree/main"
author: "Quang Mai"
date: today
date-format: long
format: pdf
editor: visual
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(knitr)
library(opendatatoronto)
library(dplyr)
library(readr)
library(kableExtra)
library(here)

@@

This paper thus inspects gender and racial identities of all Toronto police arrests from 2020 to 2021, and assesses if there is a correlation between marginalized identities and police strip searching. Foremost, this paper aims to be a form of data activism to counter the injustices related to police power abuse around strip searching of the vulnerable Toronto population. To analyze, I first map out the distribution of Toronto police arrests by race and gender and reasons for strip searching from 2020 to 2021 using bar graphs and various other graphs. Particularly, for each variable of race and gender, the distribution bars would be sectioned into 'including strip searches' or 'not including' to visualize the statistics of strip searching. With that, It's worth noting that this dataset may already contain much racial and gender bias, because it was retrieved from the very source that this paper is questioning: the Toronto Police Services [@stripSearchOpEd]. Then, the findings presented might not fully encapsulate the actual landscape of strip searching in Toronto from 2020 to 2021. Yet, it's still worth looking at due the alarming precedents of police power abuse [@covidStripSearch].

@@

This chosen dataset was published by Toronto Police Services to convey information related to all police arrests and strip searches from 2020 to 2021 (last updated on December 2, 2022). Originally, the dataset features information on date of arrest, event id, the arrested person's race, sex, and age group, arrest location, whether the arrest includes strip searching, the person's action at the time of arrest, and the police's reason for strip searching along with details on any items found. Informative as it is, the dataset contains much naming inconsistency and many variables outside the scope of this paper. Thus, in order to conduct an intersectional---or gender and race---analysis of police strip searches, I have to further clean this dataset and split it into two smaller datasets. Additional details on data extraction and cleaning will be specified in subsequent sections.

@@

#referenced kable stying from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
head(cleaned_race_gender_data, 12) |>
  kable(
    col.names = c("ID", "Race", "Gender", "Strip Search", "Items Found"),
    booktabs = TRUE
  )
```

@@

The other dataset conveys all police arrests that involve strip searching and the reasons behind strip searching. Under close examination, the naming of individual data cells in different columns from the original, raw dataset is quite disordered, for instance, using `None` and `XX` in place of the usual `NA`. Thus, all inconsistent data cells were renamed to streamline the naming conventions and data processing later on. After renaming, the big dataset was split into containing relevant columns ([see @tbl-cleaned_search_reasons_data]). Then, for this particular dataset, all instances of police arrests without strip searching are dropped using the `filter()` function, along with a `!is.na()` function for rows with `NA` as values all the `reason` column.

Except for "ID" column, all other columns in this new dataset are reasons which account for why the Toronto police force conducts strip search under certain arrests. For instance, "Reason - Assist Escape" is for people who may have helped other suspects escape and "Reason - Possess Evidence" is for people who are suspected of harboring items like drugs or an alibi. For each "Reason" column, if the data cell contains `0`, this means that strip search doesn't happen because of that particular reason. Vice versa, if it contains `1`, this means strip search happens because of that reason.

@@

head(cleaned_search_reasons_data, 12) |>
  kable(
    col.names = c("ID", "Reason - Injury", "Reason - Escape", "Reason - Weapons", "Reason - Has Evidence"),
    booktabs = TRUE
  )
```

@@

@@

```{r}
#| echo: false
#| message: false
#| label: fig-race_strip_search
#| fig-cap: Racial Identities and Police Strip Search Per Arrest in Toronto, 2020-2021

@@

@@

cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  group_by(gender) |>
  mutate(percentage = round(number_of_arrests/sum(number_of_arrests)*100)) |>
  kable(
    col.names = c("Gender", "Strip Search", "Number of Arrests", "Percentage"),
    booktabs = TRUE)
```

@@

```
### Police Strip Searches and Items Found Statistics

```{r}
#| echo: false
#| message: false
#| label: tbl-items-strip-search
#| tbl-cap: Summary Statistics of Items Found During Police Strip Searches

@@

```{r}
#| echo: false
#| include: false
#| message: false
  cleaned_arrests_strip_data |>
  select(event_id, reason_injury,	reason_escape, reason_weapons, reason_possess_evidence, items_found) |>
  filter(reason_injury == 0 & reason_escape == 0 & reason_weapons == 0 & reason_possess_evidence == 0)
```

>>>>

---
title: "Slight Positive Correlation Between Gendered and Racialized Police Strip Searches in the Toronto Municipal, 2020-2021"
title-block-banner: true
abstract: "Police strip search has been a topic of great contention during the COVID-19 pandemic. The question commonly raised was how certain demographics are more likely to undergo arrests and strip searches than others, and why has this skewed power dynamics been maintained for so long. Data then was gathered from Open Data Toronto to assess the correlation between strip searches and the racial and gender identities of arrested 'suspects' in the Toronto Municipal from 2020 to 2021. This analysis would enable us to rethink the modern form of policing, how it transpires in times of crisis and critically affects people of marginalized identities. More on RESULTS" 
thanks: "Code and data are available at: https://github.com/ponolite/Police_Arrests_and_Strip_Searches/tree/main"
author: "Quang Mai"
date: today
date-format: long
format: pdf
editor: visual
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(knitr)
library(opendatatoronto)
library(dplyr)
library(readr)
library(kableExtra)
library(here)
library(ggbeeswarm)
library(ggrepel)
library(ggplot2)

@@

This paper thus inspects gender and racial identities of all Toronto police arrests from 2020 to 2021, and assesses if there is a correlation between marginalized identities and police strip searching. Foremost, this paper aims to be a form of data activism to counter the injustices related to police power abuse around strip searching of the vulnerable Toronto population. To analyze, I first map out the distribution of Toronto police arrests by race and gender and reasons for strip searching from 2020 to 2021 using bar graphs and various other graphs. Particularly, for each category of race and gender, the distribution bars would be sectioned into 'including strip searches' or 'not including' to visualize the statistics of strip searching. With that, It's worth noting that this dataset may already contain much racial and gender bias, because it was retrieved from the very source that this paper is questioning: the Toronto Police Services [@stripSearchOpEd]. Then, the findings presented might not fully encapsulate the actual landscape of strip searching in Toronto from 2020 to 2021. Yet, it's still worth looking at due the precedents of police power abuse [@covidStripSearch].

@@

This chosen dataset was published by Toronto Police Services to convey information related to all police arrests and strip searches from 2020 to 2021 (last updated on December 2, 2022). Originally, the dataset features information on date of arrest, event id, the arrested person's race, sex, and age group, arrest location, whether the arrest includes strip searching, the person's action at the time of arrest, and the police's reason for strip searching along with details on any items found. Informative as it is, the dataset contains much naming inconsistency and many columns outside the scope of this paper. Thus, in order to conduct an intersectional---or gender and race---analysis of police strip searches, I have to further clean this dataset and split it into two smaller datasets. Additional details on data extraction and cleaning will be specified in subsequent sections.

@@

#referenced kable stying from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
head(cleaned_race_gender_data, 10) |>
  kable(
    col.names = c("ID", "Race", "Gender", "Strip Search", "Items Found"),
    booktabs = TRUE
  )
```

@@

The other dataset conveys all police arrests that involve strip searching and the reasons behind strip searching. Under close examination, the naming of individual data cells in different columns from the original, raw dataset is quite disordered, for instance, using `None` and `XX` in place of the usual `NA`. Thus, all inconsistent data cells were renamed to streamline the naming conventions and data processing later on. After renaming, the big dataset was split into containing relevant columns ([see @tbl-cleaned_search_reasons_data]). Then, for this particular dataset, all instances of police arrests without strip searching are dropped using the `filter()` function, along with a `!is.na()` function for rows with `NA` as values for all the `reason` column.

Except for "ID" column, all other columns in this new dataset are reasons which account for why the Toronto police force conducts strip search under certain arrests. For instance, "Reason - Assist Escape" is for people who may have helped other suspects escape and "Reason - Possess Evidence" is for people who are suspected of harboring items like drugs or an alibi. For each "Reason" column, if the data cell contains `0`, this means that strip search doesn't happen because of that particular reason. Vice versa, if it contains `1`, this means strip search happens because of that reason. In addition, a new column called "Total Reasons" totals all reasons used to account for each police strip search is added with `mutate()`, where all instances of `1` and `0` are changed into numeric with `as.numeric()`.

@@

  head(cleaned_search_reasons_data, 10) |>
  select(-c(race, gender)) |>
  kable(
    col.names = c("ID", "Reason - Injury", "Reason - Escape", "Reason - Weapons", "Reason - Has Evidence", "Items_Found", "Total Reasons"),
    booktabs = TRUE
  ) |>
  kable_styling(latex_options="scale_down")
```

@@

cite toronto population to say that while the number of white poeple seem large, it's not really

@@

```{r fig.height=6, fig.width=7}
#| echo: false
#| message: false
#| label: fig-race_strip_search
#| fig-cap: Racial Identities and Police Strip Search Per Arrest in Toronto, 2020-2021

@@

```{r fig.height=6, fig.width=7}
#| echo: false
#| message: false
#| label: fig-race_strip_search_separate
#| fig-cap: Racial Identities and Police Strip Search Per Arrest in Toronto, 2020-2021

cleaned_race_gender_data |>
  group_by(race, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  filter(strip_search == "Yes") |> 
  ggplot(mapping = aes(x = strip_search, fill = strip_search), data=cleaned_race_gender_data) +
  geom_bar()+theme_minimal()+ylab("number of arrests")+facet_wrap(~race)+ scale_fill_viridis_d(option="cividis")

@@

cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  group_by(gender) |>
  mutate(percentage = round(number_of_arrests/sum(number_of_arrests)*100)) |>
  kable(
    col.names = c("Gender", "Strip Search", "Number of Arrests", "Percentage"),
    booktabs = TRUE)

@@

### Police Strip Searches and Items Found Statistics

```{r}
#| echo: false
#| message: false
#| label: tbl-reasons-strip-search
#| tbl-cap: Summary Statistics of Items Found During Police Strip Searches

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-items-strip-search
#| tbl-cap: Summary Statistics of Reasons for Police Strip Searches

cleaned_search_reasons_data |>
  group_by(total_reasons, items_found) |>
  count() |>
  rename(count = n) |>
  kable(
    col.names = c("Number of Reasons", "Items_Found", "Count"),
    booktabs = TRUE
  )
```

```{r}
#| echo: false
#| message: false
#| label: fig-scatter-plot-reasons
#| fig-cap: Relationship Between Reasons for and Items Found During Per Strip Search in Toronto, 2020-2021

#create a scatterplot to show each observation in the dataset, comparing the number of reasons to the items found in each strip search
#colour the points based on the police division where the group of shootings occurred
#code for geom_point() referenced from https://github.com/TDonofrio62/Shooting-Occurrences-in-Toronto/blob/main/Shooting_Occurrences/outputs/paper/Shooting_Occurrences.Rmd
cleaned_search_reasons_data |>
  group_by(total_reasons, items_found, race) |>
  count() |>
  rename(count = n) |>
  ggplot(aes(x = total_reasons, y = count, colour = race)) +
  geom_point() +
  theme_bw() +
  theme(legend.key.size = unit(0.5, 'cm'))+
  labs(x = "Total Reasons for Strip Searching",
       y = "Count",
       color = "Race"
       ) +
  scale_fill_viridis_d(option="cividis")

cleaned_search_reasons_data |>
  group_by(total_reasons, items_found, race) |>
  count() |>
  rename(count = n) |>
  ggplot(aes(x = total_reasons, y = count, colour = race)) +
  geom_point() +
  theme_bw() +
  theme(legend.key.size = unit(0.5, 'cm'))+
  labs(x = "Total Reasons for Strip Searching",
       y = "Count",
       color = "Race"
       ) +
  scale_fill_viridis_d(option="cividis") 
```
